---
title: "DSMA Sector Proposals - Preview"
subtitle: "Departure Sectorization for EDDF, EDDM, EGLL"
date: 2026-09-02
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
---

## Overview

Application of the valley-based sectorization algorithm to departure (DEP) flows. Same methodology as validated for arrivals (ARR).

**Algorithm parameters:**
- Smoothing bandwidth: 12 degrees
- Minimum prominence: 0.02 (2%)
- Valley safety fraction: 0.25
- Timing threshold: 80 seconds

## Data Preparation

```{r}
#| label: setup
#| message: false

library(dplyr)
library(ggplot2)
library(arrow)
devtools::load_all(quiet = TRUE)

# Function to prepare one airport
prepare_airport_dep <- function(icao) {
  # Read from zip - use working directory extraction
  temp_dir <- tempdir()

  if (icao == "LGAV") {
    filename <- "LGAV_01-JAN-2024_31-DEC-2024.parquet"
    unzip(
      "../xx-test-gotcha/data/apdf-annual/apdf-LGAV-2019-2024.zip",
      files = filename,
      exdir = temp_dir, junkpaths = TRUE
    )
    apdf <- read_parquet(file.path(temp_dir, filename))
    unlink(file.path(temp_dir, filename))
  } else {
    filename <- paste0(icao, "_01-JAN-2024_31-DEC-2024.gz.parquet")
    unzip(
      "../xx-test-gotcha/data/apdf-annual/apdf-2024.zip",
      files = filename,
      exdir = temp_dir, junkpaths = TRUE
    )
    apdf <- read_parquet(file.path(temp_dir, filename))
    unlink(file.path(temp_dir, filename))
  }

  # Prepare TMA samples
  tma <- apdf |>
    filter(SRC_PHASE == "DEP") |>
    filter(!is.na(C40_BEARING) | !is.na(C100_BEARING)) |>
    tidyr::pivot_longer(
      cols = c(C40_BEARING, C100_BEARING),
      names_to = "RANGE_TYPE",
      values_to = "BEARING"
    ) |>
    mutate(
      RANGE_NM = if_else(RANGE_TYPE == "C40_BEARING", 40, 100),
      TMA_ADDL_TIME_MIN = if_else(
        RANGE_NM == 40,
        C40_TRANSIT_TIME_MIN,
        C100_TRANSIT_TIME_MIN
      ),
      PHASE = SRC_PHASE,
      DATE = as.Date(MVT_TIME_UTC),
      RWY = AP_C_RWY,
      FLIGHT_ID = AP_C_FLTID,
      ICAO = icao
    ) |>
    filter(
      !is.na(BEARING),
      BEARING >= 0,
      BEARING < 360,
      !is.na(TMA_ADDL_TIME_MIN),
      TMA_ADDL_TIME_MIN > 0
    ) |>
    select(ICAO, PHASE, DATE, RANGE_NM, BEARING, TMA_ADDL_TIME_MIN,
           RWY, AC_CLASS, FLIGHT_ID)

  return(tma)
}

# Process airports
cat("Loading airports...\n")
tma_eddf <- prepare_airport_dep("EDDF")
tma_eddm <- prepare_airport_dep("EDDM")
tma_egll <- prepare_airport_dep("EGLL")

tma_all <- bind_rows(tma_eddf, tma_eddm, tma_egll)

cat("Loaded", format(nrow(tma_all), big.mark = ","), "DEP movements\n")
```

**Sample sizes:**
```{r}
#| label: sample-summary

tma_all |>
  summarise(Movements = n(), .by = c(ICAO, RANGE_NM)) |>
  arrange(ICAO, RANGE_NM) |>
  knitr::kable()
```

## EDDF - Frankfurt

```{r}
#| label: eddf-40nm
#| fig-width: 11
#| fig-height: 5

d <- prepare_tma_bearing_density(tma_all, "EDDF", "DEP", 40, 12)
e <- identify_tma_bearing_extrema(d, 0.02)
p <- propose_tma_sector_definitions(d, e, 5, 0.25)

ggplot(d, aes(x = BEARING_MID, y = SMOOTHED_N)) +
  geom_line(colour = "#2b6a8a", linewidth = 0.8) +
  geom_vline(data = p, aes(xintercept = BEARING_FROM),
             colour = "#c8553d", linewidth = 0.8) +
  scale_x_continuous(breaks = seq(0, 360, by = 30)) +
  coord_cartesian(xlim = c(0, 360), expand = FALSE) +
  labs(
    title = sprintf("EDDF DEP 40 NM - %d Sectors Proposed", nrow(p)),
    x = "Bearing (degrees)", y = "Smoothed movements per degree"
  ) +
  theme_minimal(base_size = 11)
```

```{r}
#| label: eddf-100nm
#| fig-width: 11
#| fig-height: 5

d <- prepare_tma_bearing_density(tma_all, "EDDF", "DEP", 100, 12)
e <- identify_tma_bearing_extrema(d, 0.02)
p <- propose_tma_sector_definitions(d, e, 5, 0.25)

ggplot(d, aes(x = BEARING_MID, y = SMOOTHED_N)) +
  geom_line(colour = "#2b6a8a", linewidth = 0.8) +
  geom_vline(data = p, aes(xintercept = BEARING_FROM),
             colour = "#c8553d", linewidth = 0.8) +
  scale_x_continuous(breaks = seq(0, 360, by = 30)) +
  coord_cartesian(xlim = c(0, 360), expand = FALSE) +
  labs(
    title = sprintf("EDDF DEP 100 NM - %d Sectors Proposed", nrow(p)),
    x = "Bearing (degrees)", y = "Smoothed movements per degree"
  ) +
  theme_minimal(base_size = 11)
```

## EDDM - Munich

```{r}
#| label: eddm-40nm
#| fig-width: 11
#| fig-height: 5

d <- prepare_tma_bearing_density(tma_all, "EDDM", "DEP", 40, 12)
e <- identify_tma_bearing_extrema(d, 0.02)
p <- propose_tma_sector_definitions(d, e, 5, 0.25)

ggplot(d, aes(x = BEARING_MID, y = SMOOTHED_N)) +
  geom_line(colour = "#2b6a8a", linewidth = 0.8) +
  geom_vline(data = p, aes(xintercept = BEARING_FROM),
             colour = "#c8553d", linewidth = 0.8) +
  scale_x_continuous(breaks = seq(0, 360, by = 30)) +
  coord_cartesian(xlim = c(0, 360), expand = FALSE) +
  labs(
    title = sprintf("EDDM DEP 40 NM - %d Sectors Proposed", nrow(p)),
    x = "Bearing (degrees)", y = "Smoothed movements per degree"
  ) +
  theme_minimal(base_size = 11)
```

```{r}
#| label: eddm-100nm
#| fig-width: 11
#| fig-height: 5

d <- prepare_tma_bearing_density(tma_all, "EDDM", "DEP", 100, 12)
e <- identify_tma_bearing_extrema(d, 0.02)
p <- propose_tma_sector_definitions(d, e, 5, 0.25)

ggplot(d, aes(x = BEARING_MID, y = SMOOTHED_N)) +
  geom_line(colour = "#2b6a8a", linewidth = 0.8) +
  geom_vline(data = p, aes(xintercept = BEARING_FROM),
             colour = "#c8553d", linewidth = 0.8) +
  scale_x_continuous(breaks = seq(0, 360, by = 30)) +
  coord_cartesian(xlim = c(0, 360), expand = FALSE) +
  labs(
    title = sprintf("EDDM DEP 100 NM - %d Sectors Proposed", nrow(p)),
    x = "Bearing (degrees)", y = "Smoothed movements per degree"
  ) +
  theme_minimal(base_size = 11)
```

## EGLL - London Heathrow

```{r}
#| label: egll-40nm
#| fig-width: 11
#| fig-height: 5

d <- prepare_tma_bearing_density(tma_all, "EGLL", "DEP", 40, 12)
e <- identify_tma_bearing_extrema(d, 0.02)
p <- propose_tma_sector_definitions(d, e, 5, 0.25)

ggplot(d, aes(x = BEARING_MID, y = SMOOTHED_N)) +
  geom_line(colour = "#2b6a8a", linewidth = 0.8) +
  geom_vline(data = p, aes(xintercept = BEARING_FROM),
             colour = "#c8553d", linewidth = 0.8) +
  scale_x_continuous(breaks = seq(0, 360, by = 30)) +
  coord_cartesian(xlim = c(0, 360), expand = FALSE) +
  labs(
    title = sprintf("EGLL DEP 40 NM - %d Sectors Proposed", nrow(p)),
    x = "Bearing (degrees)", y = "Smoothed movements per degree"
  ) +
  theme_minimal(base_size = 11)
```

```{r}
#| label: egll-100nm
#| fig-width: 11
#| fig-height: 5

d <- prepare_tma_bearing_density(tma_all, "EGLL", "DEP", 100, 12)
e <- identify_tma_bearing_extrema(d, 0.02)
p <- propose_tma_sector_definitions(d, e, 5, 0.25)

ggplot(d, aes(x = BEARING_MID, y = SMOOTHED_N)) +
  geom_line(colour = "#2b6a8a", linewidth = 0.8) +
  geom_vline(data = p, aes(xintercept = BEARING_FROM),
             colour = "#c8553d", linewidth = 0.8) +
  scale_x_continuous(breaks = seq(0, 360, by = 30)) +
  coord_cartesian(xlim = c(0, 360), expand = FALSE) +
  labs(
    title = sprintf("EGLL DEP 100 NM - %d Sectors Proposed", nrow(p)),
    x = "Bearing (degrees)", y = "Smoothed movements per degree"
  ) +
  theme_minimal(base_size = 11)
```

## Summary

All three airports show clear departure flow structure with 5-6 sectors per range, consistent with the 4-7 sector target validated for arrivals.
