---
title: "ASMA/DSMA development"
format: html
---

## Problem Space

By tradition we are using a "histogram" based approach to identify entry sectors for ASMA.   
This approach shall be also applied to DSMA, i.e., identifying respective exit sectors.

Using "sectors" offers flexibility as changes to the operational environment and airspace organisation will undergo changes.
Entry/exit fixes or waypoints may be moved, introduced/removed.   
Experience within the European airspace has demonstrated that such "system point" changes occur regularly making it impossible to maintain a respective database (in sync with AIRAC cycles).    
Tactical sequencing and selective use of nuances in the clearance limit, make flights pass abeam such waypoints.

Sector identification is based on the crossing of the respective approximation of the arrival/depature airspace.    
By definition we use concentric circles of radius 40NM and 100NM around the aerodrome reference point (ARP).    
Note: the vertical flight efficiency (VFE) horizon is 200NM. With a view to study XSMA/VFE, we may need to extend the ASMA/DSMA = XSMA references to also include 200NM ranges.    
Conceptually, the algorithm should be paramterised, if future studies would require/identify another radius.

The visual sector identification aims at identifying up to about 5-6 "sectors" unless there is operational evidence that the operations can be better described by using additional sectors.

The XSMA algorithm determines a reference time for similar flight operations.   
Similar flight operations are characterised by the 
* respective entry/exit sector;
* aircraft velocity range as a combination of weight turbulence category and propulsion system, i.e. L/M/H & P/T/J (e.g. Light Piston LP, Medium Turboprop MT, Heavy Jet HJ), while the A380 and a Antonov are typically associated as "Super" heavy 'S' from a WTC perspective, these aircraft as grouped together with HJ.
* runway used for landing/ASMA or depature/DSMA.

Discussion are on-going about adapting the newer groupings based on the recent recatorigation RECAT of aircraft in a pair-to-pair relationship to optimise the pairwise WTC separation.    
This is currently out of scope, but from a conceptual perspective it should not influence the sectorisation identification. 
The newer RECAT labels being a further breakdown of the existing WTC.   
It may however influence sample sizes required for the determination of the reference times.

The sample size for the determination of reference times is currently set as n = 20, given that the reference period covers movement within a full year.


## Approach

The XSMA analysis builds on the establishment of a reference time sample for the respective study airport.

For this purpose we work with the 40NM and 100NM milestones embedded in the APDF data set.
The APDF data set also provides for the bearing from the aerodrome reference point to the flight milestone, i.e., we do not have to calculate this from the ARP.

### Reference Period Governance

Each study establishes an agreed start reference set for its selected reference
period. The 2024 set developed here is therefore specific to the current study;
other studies may agree a different reference period. Once established, the
study reference remains in use for subsequent analysis batches until an
operational change or a new study decision requires a revised sectorisation and
reference set.


## Example - EGLL and LSZH

We develop the initial sector-identification logic using EGLL and LSZH. The
first diagnostic is the observed distribution of arrival crossing bearings.
For this purpose, it counts every known APDF crossing bearing, rather than only
movements that already pass the later class, runway, and time-validity filters.
This keeps the sector proposal anchored in the geometry of the traffic flow.

```{r}
#| label: setup
#| include: false

library(dplyr)
library(ggplot2)
devtools::load_all("..", quiet = TRUE)
```

### EGLL arrivals, 2024

```{r}
#| label: egll-arrival-bearings
#| include: false

airport <- "EGLL"
phase <- "ARR"
reference_year <- 2024L
bearing_bin_width <- 6

apdf_archive <- Sys.getenv(
  "PBWG_APDF_2024",
  unset = "/Users/rainerkoelle/RProjects/__DATA/APDF/apdf-2024.zip"
)
egll_file <- "EGLL_01-JAN-2024_31-DEC-2024.gz.parquet"

egll_arrival_samples <- prepare_apdf_tma_reference_input_from_zip(
  zipped_archive_path = apdf_archive,
  files = egll_file,
  type = "parquet",
  ranges = c(40, 100)
)

egll_arrival_histogram <- prepare_tma_bearing_histogram(
  tma_samples = egll_arrival_samples,
  airport = airport,
  phase = phase,
  ranges = c(40, 100),
  bearing_bin_width = bearing_bin_width
) |>
  mutate(RANGE_NM = factor(RANGE_NM, levels = c(100, 40)))
```

The default bin width is 6 degrees, producing 60 equal bins around the ARP.
`bearing_bin_width` is deliberately configurable, but must divide 360 exactly.
The 100 NM horizon is shown above the 40 NM horizon so the two distributions
can be compared on the same bearing scale. To render this note outside the
development environment, set `PBWG_APDF_2024` to the local 2024 APDF archive.

```{r}
#| label: egll-arrival-bearing-histogram
#| fig-width: 11
#| fig-height: 7
#| fig-cap: "EGLL arrival crossing bearings in 2024. Each bar is a 6 degree bearing bin; all known APDF crossing bearings are included."

ggplot(
  egll_arrival_histogram,
  aes(x = .data$BEARING_MID, y = .data$N)
) +
  geom_col(width = bearing_bin_width, fill = "#2b6a8a", colour = "white", linewidth = 0.15) +
  facet_wrap(~RANGE_NM, ncol = 1, scales = "free_y", labeller = label_both) +
  scale_x_continuous(breaks = seq(0, 360, by = 30)) +
  coord_cartesian(xlim = c(0, 360), expand = FALSE) +
  labs(
    title = paste(airport, phase, "crossing bearings", reference_year),
    subtitle = paste("Known APDF bearings;", bearing_bin_width, "degree bins"),
    x = "Bearing from aerodrome reference point (degrees)",
    y = "Movements"
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```

### Smoothed Density And Local Extrema

The one-degree counts are retained as the analytical grid, but not interpreted
as one-degree operational detail. A wrapped Gaussian smoother reduces local
noise while treating 360 and 000 degrees as adjacent. The default bandwidth is
6 degrees, with an effective influence of approximately 18 degrees to either
side. The prominence threshold only identifies substantial candidate extrema;
it does not yet approve a sector boundary.

```{r}
#| label: egll-arrival-bearing-density
#| include: false

smoothing_bandwidth <- 6
minimum_relative_prominence <- 0.02

egll_arrival_density <- prepare_tma_bearing_density(
  tma_samples = egll_arrival_samples,
  airport = airport,
  phase = phase,
  ranges = c(40, 100),
  smoothing_bandwidth = smoothing_bandwidth
) |>
  mutate(RANGE_NM = factor(RANGE_NM, levels = c(100, 40)))

egll_arrival_extrema <- identify_tma_bearing_extrema(
  tma_density = egll_arrival_density,
  min_relative_prominence = minimum_relative_prominence
) |>
  mutate(RANGE_NM = factor(RANGE_NM, levels = c(100, 40)))
```

```{r}
#| label: egll-arrival-bearing-density-plot
#| fig-width: 11
#| fig-height: 7
#| fig-cap: "EGLL arrival bearing density in 2024. Triangles mark local peaks and inverted triangles local minima; filled markers pass the 2 percent relative-prominence screen. Labels identify substantial peaks only."

ggplot(
  egll_arrival_density,
  aes(x = .data$BEARING_MID, y = .data$SMOOTHED_N)
) +
  geom_line(colour = "#2b6a8a", linewidth = 0.8) +
  geom_point(
    data = egll_arrival_extrema,
    aes(
      x = .data$BEARING,
      y = .data$SMOOTHED_N,
      shape = .data$EXTREMUM,
      fill = .data$IS_SUBSTANTIAL
    ),
    size = 2.8,
    colour = "#263238",
    inherit.aes = FALSE
  ) +
  geom_text(
    data = filter(egll_arrival_extrema, .data$EXTREMUM == "PEAK", .data$IS_SUBSTANTIAL),
    aes(
      x = .data$BEARING,
      y = .data$SMOOTHED_N,
      label = sprintf("%.0f", .data$BEARING)
    ),
    vjust = 1.7,
    size = 3,
    inherit.aes = FALSE
  ) +
  facet_wrap(~RANGE_NM, ncol = 1, scales = "free_y", labeller = label_both) +
  scale_x_continuous(breaks = seq(0, 360, by = 30)) +
  scale_shape_manual(values = c(PEAK = 24, MINIMUM = 25)) +
  scale_fill_manual(values = c(`TRUE` = "#f4a261", `FALSE` = "white")) +
  guides(fill = "none") +
  coord_cartesian(xlim = c(0, 360), expand = FALSE) +
  labs(
    title = paste(airport, phase, "smoothed crossing-bearing density", reference_year),
    subtitle = paste("One-degree counts; circular Gaussian bandwidth", smoothing_bandwidth, "degrees"),
    x = "Bearing from aerodrome reference point (degrees)",
    y = "Smoothed movements per degree",
    shape = "Local extremum"
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```

For a peak, relative prominence is calculated as:

```text
(peak height - higher adjacent minimum) / highest density in the range
```

For a minimum, relative valley depth is calculated as:

```text
(lower adjacent peak - minimum height) / highest density in the range
```

```{r}
#| label: egll-arrival-extrema-table

egll_arrival_extrema |>
  filter(.data$IS_SUBSTANTIAL) |>
  transmute(
    `Range (NM)` = .data$RANGE_NM,
    Type = .data$EXTREMUM,
    `Bearing (degrees)` = round(.data$BEARING, 1),
    `Smoothed movements` = round(.data$SMOOTHED_N),
    Metric = if_else(
      .data$EXTREMUM == "PEAK",
      "Relative peak prominence",
      "Relative valley depth"
    ),
    Value = scales::percent(.data$RELATIVE_PROMINENCE, accuracy = 0.1)
  )
```

### Proposed Sector Boundaries

Candidate cuts are the lowest-density positions between substantial peaks. A
candidate is rounded to the nearest 5-degree boundary only when that position
remains within the lower 25 percent of the valley rise to its smaller adjacent
peak. Where North lies safely in the final valley, it becomes the explicit
`000` seam rather than creating a North-wrapping sector.

```{r}
#| label: egll-arrival-sector-proposal
#| include: false

rounding_increment <- 5
valley_safety_fraction <- 0.25

egll_arrival_proposal <- propose_tma_sector_definitions(
  tma_density = egll_arrival_density,
  extrema = egll_arrival_extrema,
  rounding_increment = rounding_increment,
  valley_safety_fraction = valley_safety_fraction
)

egll_arrival_sector_definitions <- egll_arrival_proposal$sector_definitions |>
  mutate(RANGE_NM = factor(RANGE_NM, levels = c(100, 40)))

egll_arrival_cut_audit <- egll_arrival_proposal$cut_audit |>
  mutate(RANGE_NM = factor(RANGE_NM, levels = c(100, 40)))

egll_arrival_sector_support <- summarise_tma_sector_support(
  tma_samples = egll_arrival_samples,
  sector_definitions = egll_arrival_proposal$sector_definitions
)
```

```{r}
#| label: egll-arrival-proposed-boundaries
#| fig-width: 11
#| fig-height: 7
#| fig-cap: "EGLL proposed 2024 ARR sector boundaries. Solid vertical lines are valley-safe rounded boundaries; the 000 line is the non-wrapping North seam."

ggplot(
  egll_arrival_density,
  aes(x = .data$BEARING_MID, y = .data$SMOOTHED_N)
) +
  geom_line(colour = "#2b6a8a", linewidth = 0.8) +
  geom_vline(
    data = egll_arrival_cut_audit,
    aes(xintercept = .data$CUT_ROUNDED),
    colour = "#c8553d",
    linewidth = 0.7,
    linetype = "dashed",
    inherit.aes = FALSE
  ) +
  facet_wrap(~RANGE_NM, ncol = 1, scales = "free_y", labeller = label_both) +
  scale_x_continuous(breaks = seq(0, 360, by = 30)) +
  coord_cartesian(xlim = c(0, 360), expand = FALSE) +
  labs(
    title = paste(airport, phase, "proposed sector boundaries", reference_year),
    subtitle = paste("Valley-safe", rounding_increment, "degree rounding"),
    x = "Bearing from aerodrome reference point (degrees)",
    y = "Smoothed movements per degree"
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```

```{r}
#| label: egll-arrival-cut-audit

egll_arrival_cut_audit |>
  transmute(
    `Range (NM)` = .data$RANGE_NM,
    `Left peak` = round(.data$LEFT_PEAK, 1),
    `Right peak` = round(.data$RIGHT_PEAK, 1),
    `Raw cut` = round(.data$CUT_RAW, 1),
    `Final cut` = .data$CUT_ROUNDED,
    `Rounding delta` = round(.data$ROUNDING_DELTA, 1),
    Status = .data$ROUNDING_STATUS,
    `North overrun` = .data$NORTH_OVERRUN
  )
```

### Movement And Reference-Cell Support

The candidate sectors are now assessed without applying a sample-size
threshold. The table shows traffic volume and the distribution of 2024 valid
reference-cell counts by runway and aircraft class. `SECTOR_ID` is the fully
contextual identifier retained for audit and maintenance; `SECTOR` remains the
compact bearing label used alongside the ICAO and phase columns in the lookup.

```{r}
#| label: egll-arrival-sector-support

egll_arrival_sector_support$sector_summary |>
  transmute(
    `Range (NM)` = .data$RANGE_NM,
    `Sector ID` = .data$SECTOR_ID,
    `Crossings` = .data$N_CROSSINGS,
    `Range share` = scales::percent(.data$PCT_RANGE_CROSSINGS, accuracy = 0.1),
    `Valid TMA` = .data$N_VALID_TMA,
    `Reference cells` = .data$N_REFERENCE_CELLS,
    `Min cell n` = .data$MIN_CELL_N,
    `P25 cell n` = round(.data$P25_CELL_N),
    `Median cell n` = round(.data$MEDIAN_CELL_N),
    `Max cell n` = .data$MAX_CELL_N
  )
```

### LSZH Arrivals, 2024

LSZH applies the same diagnostic settings as EGLL. Its intersecting-runway
operation provides a contrasting test: the method may legitimately identify a
different number of flows at 40 NM and 100 NM, and a sector may cross North
when the observed bearing density is continuous across 360 and 000 degrees.

```{r}
#| label: lszh-arrival-input
#| include: false

lszh_file <- "LSZH_01-JAN-2024_31-DEC-2024.gz.parquet"

lszh_arrival_samples <- prepare_apdf_tma_reference_input_from_zip(
  zipped_archive_path = apdf_archive,
  files = lszh_file,
  type = "parquet",
  ranges = c(40, 100)
)

lszh_arrival_histogram <- prepare_tma_bearing_histogram(
  tma_samples = lszh_arrival_samples,
  airport = "LSZH",
  phase = "ARR",
  ranges = c(40, 100),
  bearing_bin_width = bearing_bin_width
) |>
  mutate(RANGE_NM = factor(RANGE_NM, levels = c(100, 40)))

lszh_arrival_density <- prepare_tma_bearing_density(
  tma_samples = lszh_arrival_samples,
  airport = "LSZH",
  phase = "ARR",
  ranges = c(40, 100),
  smoothing_bandwidth = smoothing_bandwidth
) |>
  mutate(RANGE_NM = factor(RANGE_NM, levels = c(100, 40)))

lszh_arrival_extrema <- identify_tma_bearing_extrema(
  tma_density = lszh_arrival_density,
  min_relative_prominence = minimum_relative_prominence
) |>
  mutate(RANGE_NM = factor(RANGE_NM, levels = c(100, 40)))

lszh_arrival_proposal <- propose_tma_sector_definitions(
  tma_density = lszh_arrival_density,
  extrema = lszh_arrival_extrema,
  rounding_increment = rounding_increment,
  valley_safety_fraction = valley_safety_fraction
)

lszh_arrival_cut_audit <- lszh_arrival_proposal$cut_audit |>
  mutate(RANGE_NM = factor(RANGE_NM, levels = c(100, 40)))

lszh_arrival_sector_support <- summarise_tma_sector_support(
  tma_samples = lszh_arrival_samples,
  sector_definitions = lszh_arrival_proposal$sector_definitions
)
```

```{r}
#| label: lszh-arrival-bearing-histogram
#| fig-width: 11
#| fig-height: 7
#| fig-cap: "LSZH arrival crossing bearings in 2024. Each bar is a 6 degree bearing bin; all known APDF crossing bearings are included."

ggplot(
  lszh_arrival_histogram,
  aes(x = .data$BEARING_MID, y = .data$N)
) +
  geom_col(width = bearing_bin_width, fill = "#2b6a8a", colour = "white", linewidth = 0.15) +
  facet_wrap(~RANGE_NM, ncol = 1, scales = "free_y", labeller = label_both) +
  scale_x_continuous(breaks = seq(0, 360, by = 30)) +
  coord_cartesian(xlim = c(0, 360), expand = FALSE) +
  labs(
    title = "LSZH ARR crossing bearings 2024",
    subtitle = paste("Known APDF bearings;", bearing_bin_width, "degree bins"),
    x = "Bearing from aerodrome reference point (degrees)",
    y = "Movements"
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```

```{r}
#| label: lszh-arrival-proposed-boundaries
#| fig-width: 11
#| fig-height: 7
#| fig-cap: "LSZH proposed 2024 ARR sector boundaries. Dashed lines are valley-safe rounded boundaries; the final sector crosses North where the density supports it."

ggplot(
  lszh_arrival_density,
  aes(x = .data$BEARING_MID, y = .data$SMOOTHED_N)
) +
  geom_line(colour = "#2b6a8a", linewidth = 0.8) +
  geom_vline(
    data = lszh_arrival_cut_audit,
    aes(xintercept = .data$CUT_ROUNDED),
    colour = "#c8553d",
    linewidth = 0.7,
    linetype = "dashed",
    inherit.aes = FALSE
  ) +
  facet_wrap(~RANGE_NM, ncol = 1, scales = "free_y", labeller = label_both) +
  scale_x_continuous(breaks = seq(0, 360, by = 30)) +
  coord_cartesian(xlim = c(0, 360), expand = FALSE) +
  labs(
    title = "LSZH ARR proposed sector boundaries 2024",
    subtitle = paste("Valley-safe", rounding_increment, "degree rounding"),
    x = "Bearing from aerodrome reference point (degrees)",
    y = "Smoothed movements per degree"
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```

```{r}
#| label: lszh-arrival-cut-audit

lszh_arrival_cut_audit |>
  transmute(
    `Range (NM)` = .data$RANGE_NM,
    `Left peak` = round(.data$LEFT_PEAK, 1),
    `Right peak` = round(.data$RIGHT_PEAK, 1),
    `Raw cut` = round(.data$CUT_RAW, 1),
    `Final cut` = .data$CUT_ROUNDED,
    `Rounding delta` = round(.data$ROUNDING_DELTA, 1),
    Status = .data$ROUNDING_STATUS,
    `North overrun` = .data$NORTH_OVERRUN
  )
```

```{r}
#| label: lszh-arrival-sector-support

lszh_arrival_sector_support$sector_summary |>
  transmute(
    `Range (NM)` = .data$RANGE_NM,
    `Sector ID` = .data$SECTOR_ID,
    `Crossings` = .data$N_CROSSINGS,
    `Range share` = scales::percent(.data$PCT_RANGE_CROSSINGS, accuracy = 0.1),
    `Valid TMA` = .data$N_VALID_TMA,
    `Reference cells` = .data$N_REFERENCE_CELLS,
    `Min cell n` = .data$MIN_CELL_N,
    `P25 cell n` = round(.data$P25_CELL_N),
    `Median cell n` = round(.data$MEDIAN_CELL_N),
    `Max cell n` = .data$MAX_CELL_N
  )
```

### LSZH Multi-Peak Assessment

Close peaks are not automatically separate sectors. For each candidate cut, we
compare the valley with the lower of its adjacent peaks, repeat the proposal at
4, 6, and 8 degree smoothing bandwidths, and compare conditional 20th
percentile reference times and conditional median times for runway/class cells
shared on both sides. The 20th-percentile comparison tests the possible impact
on the GANP reference; the median comparison describes typical operational
behaviour. The `0.75` pairwise-separation value below is an investigation flag,
not an acceptance threshold.

```{r}
#| label: lszh-multipeak-input
#| include: false

analysis_bandwidths <- c(4, 6, 8)

lszh_multiscale_results <- purrr::map(
  analysis_bandwidths,
  function(bandwidth) {
    density <- prepare_tma_bearing_density(
      tma_samples = lszh_arrival_samples,
      airport = "LSZH",
      phase = "ARR",
      ranges = c(40, 100),
      smoothing_bandwidth = bandwidth
    )
    extrema <- identify_tma_bearing_extrema(
      tma_density = density,
      min_relative_prominence = minimum_relative_prominence
    )
    proposal <- propose_tma_sector_definitions(
      tma_density = density,
      extrema = extrema,
      rounding_increment = rounding_increment,
      valley_safety_fraction = valley_safety_fraction
    )

    list(density = density, cut_audit = proposal$cut_audit)
  }
)

lszh_multiscale_density <- purrr::imap_dfr(
  lszh_multiscale_results,
  function(result, index) {
    mutate(result$density, SMOOTHING_BANDWIDTH = analysis_bandwidths[[index]])
  }
)

lszh_multiscale_cuts <- purrr::imap_dfr(
  lszh_multiscale_results,
  function(result, index) {
    mutate(result$cut_audit, SMOOTHING_BANDWIDTH = analysis_bandwidths[[index]])
  }
)

circular_distance <- function(first, second) {
  abs(((first - second + 180) %% 360) - 180)
}

lszh_cut_stability <- lszh_multiscale_cuts |>
  filter(.data$SMOOTHING_BANDWIDTH == smoothing_bandwidth) |>
  transmute(
    RANGE_NM = .data$RANGE_NM,
    CUT = .data$CUT_ROUNDED,
    N_STABLE_BANDWIDTHS = purrr::map2_int(
      .data$RANGE_NM,
      .data$CUT_ROUNDED,
      function(range_nm, cut) {
        sum(purrr::map_lgl(
          analysis_bandwidths,
          function(bandwidth) {
            candidates <- lszh_multiscale_cuts |>
              filter(
                .data$RANGE_NM == range_nm,
                .data$SMOOTHING_BANDWIDTH == bandwidth
              ) |>
              pull(.data$CUT_ROUNDED)
            any(circular_distance(candidates, cut) <= rounding_increment)
          }
        ))
      }
    )
  )

lszh_cell_references <- lszh_arrival_sector_support$assigned_samples |>
  filter(.data$VALID_TMA, !is.na(.data$SECTOR)) |>
  summarise(
    N = dplyr::n(),
    GANP_P20 = as.numeric(stats::quantile(.data$TMA_TIME, probs = 0.20, names = FALSE)),
    TYPICAL_P50 = stats::median(.data$TMA_TIME),
    .by = c("RANGE_NM", "SECTOR", "RWY", "CLASS")
  )

lszh_boundary_pairs <- lszh_arrival_proposal$cut_audit |>
  left_join(
    lszh_arrival_proposal$sector_definitions |>
      transmute(
        RANGE_NM = .data$RANGE_NM,
        CUT_ROUNDED = .data$BEARING_TO,
        LEFT_SECTOR = .data$SECTOR
      ),
    by = c("RANGE_NM", "CUT_ROUNDED")
  ) |>
  transmute(
    RANGE_NM = .data$RANGE_NM,
    CUT = .data$CUT_ROUNDED,
    PEAK_SEPARATION_DEG = .data$PEAK_SEPARATION_DEG,
    VALLEY_TO_LOWER_PEAK_RATIO = .data$VALLEY_TO_LOWER_PEAK_RATIO,
    PAIRWISE_SEPARATION = .data$PAIRWISE_SEPARATION,
    LEFT_SECTOR = .data$LEFT_SECTOR,
    RIGHT_SECTOR = .data$SECTOR
  )

lszh_boundary_utility <- lszh_boundary_pairs |>
  left_join(
    lszh_cell_references |>
      rename(
        LEFT_SECTOR = .data$SECTOR,
        N_LEFT = .data$N,
        P20_LEFT = .data$GANP_P20,
        P50_LEFT = .data$TYPICAL_P50
      ),
    by = c("RANGE_NM", "LEFT_SECTOR")
  ) |>
  inner_join(
    lszh_cell_references |>
      rename(
        RIGHT_SECTOR = .data$SECTOR,
        N_RIGHT = .data$N,
        P20_RIGHT = .data$GANP_P20,
        P50_RIGHT = .data$TYPICAL_P50
      ),
    by = c("RANGE_NM", "RIGHT_SECTOR", "RWY", "CLASS")
  ) |>
  summarise(
    N_SHARED_CELLS = dplyr::n(),
    MIN_SHARED_CELL_N = min(pmin(.data$N_LEFT, .data$N_RIGHT)),
    MEDIAN_ABS_P20_DIFF = stats::median(abs(.data$P20_LEFT - .data$P20_RIGHT)),
    WEIGHTED_ABS_P20_DIFF = stats::weighted.mean(
      abs(.data$P20_LEFT - .data$P20_RIGHT),
      w = pmin(.data$N_LEFT, .data$N_RIGHT)
    ),
    MEDIAN_ABS_P50_DIFF = stats::median(abs(.data$P50_LEFT - .data$P50_RIGHT)),
    WEIGHTED_ABS_P50_DIFF = stats::weighted.mean(
      abs(.data$P50_LEFT - .data$P50_RIGHT),
      w = pmin(.data$N_LEFT, .data$N_RIGHT)
    ),
    .by = c(
      "RANGE_NM", "CUT", "PEAK_SEPARATION_DEG",
      "VALLEY_TO_LOWER_PEAK_RATIO", "PAIRWISE_SEPARATION"
    )
  ) |>
  left_join(lszh_cut_stability, by = c("RANGE_NM", "CUT")) |>
  mutate(
    GEOMETRY_REVIEW = case_when(
      .data$PAIRWISE_SEPARATION < 0.75 ~ "Shallow valley review",
      .data$PEAK_SEPARATION_DEG < 3 * smoothing_bandwidth ~ "Close peaks review",
      .default = "No geometry warning"
    )
  ) |>
  arrange(.data$RANGE_NM, .data$CUT)
```

```{r}
#| label: lszh-100nm-multiscale-density
#| fig-width: 11
#| fig-height: 8
#| fig-cap: "LSZH 100 NM arrival density at three smoothing bandwidths. The 4 degree view identifies additional small features that do not persist at 6 and 8 degrees."

ggplot(
  filter(lszh_multiscale_density, .data$RANGE_NM == 100),
  aes(x = .data$BEARING_MID, y = .data$SMOOTHED_N)
) +
  geom_line(colour = "#2b6a8a", linewidth = 0.8) +
  geom_vline(
    data = filter(lszh_multiscale_cuts, .data$RANGE_NM == 100),
    aes(xintercept = .data$CUT_ROUNDED),
    colour = "#c8553d",
    linewidth = 0.65,
    linetype = "dashed",
    inherit.aes = FALSE
  ) +
  facet_wrap(~SMOOTHING_BANDWIDTH, ncol = 1, scales = "free_y") +
  scale_x_continuous(breaks = seq(0, 360, by = 30)) +
  coord_cartesian(xlim = c(0, 360), expand = FALSE) +
  labs(
    title = "LSZH ARR 100 NM multi-scale density",
    x = "Bearing from aerodrome reference point (degrees)",
    y = "Smoothed movements per degree"
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```

```{r}
#| label: lszh-100nm-boundary-assessment

lszh_boundary_utility |>
  filter(.data$RANGE_NM == 100) |>
  transmute(
    `Cut` = .data$CUT,
    `Peak gap (deg)` = round(.data$PEAK_SEPARATION_DEG, 1),
    `Valley / lower peak` = scales::percent(.data$VALLEY_TO_LOWER_PEAK_RATIO, accuracy = 0.1),
    `Pairwise separation` = scales::percent(.data$PAIRWISE_SEPARATION, accuracy = 0.1),
    `Stable bandwidths` = .data$N_STABLE_BANDWIDTHS,
    `Shared cells` = .data$N_SHARED_CELLS,
    `Min shared n` = .data$MIN_SHARED_CELL_N,
    `Weighted |P20 diff| (min)` = round(.data$WEIGHTED_ABS_P20_DIFF, 2),
    `Weighted |P50 diff| (min)` = round(.data$WEIGHTED_ABS_P50_DIFF, 2),
    `Geometry review` = .data$GEOMETRY_REVIEW
  )
```

## Full Study-Airport ARR Review

The same proposal workflow is now applied to the complete 2024 ARR study set.
The APDF archive currently contains extracts for 11 of the 13 agreed airports:
`EDDF`, `EDDM`, `EGKK`, `EGLL`, `EHAM`, `LEBL`, `LEMD`, `LFPG`, `LIRF`, `LSZH`,
and `LTFM`. `LGAV` and `LPPT` remain in the study register, but their 2024 APDF
extracts are not present in this archive and therefore cannot yet receive a
proposal.

The density matrix is a proposal-for-approval, not a final sector definition.
Every dashed line is a valley-safe rounded candidate boundary. The decision
register identifies boundaries requiring an analyst decision because they have
a shallow valley, close peaks, or do not persist across the 4, 6, and 8 degree
smoothers. Conditional weighted P20 and P50 differences are supporting
evidence only; no materiality tolerance is applied at this stage.

```{r}
#| label: full-study-arrival-input
#| include: false

study_airports <- c(
  "EDDF", "EDDM", "EGKK", "EGLL", "EHAM", "LEBL", "LEMD",
  "LFPG", "LGAV", "LIRF", "LPPT", "LSZH", "LTFM"
)
available_study_airports <- setdiff(study_airports, c("LGAV", "LPPT"))

circular_distance <- function(first, second) {
  abs(((first - second + 180) %% 360) - 180)
}

build_boundary_utility <- function(samples, proposal, airport) {
  assigned_samples <- summarise_tma_sector_support(
    tma_samples = samples,
    sector_definitions = proposal$sector_definitions
  )$assigned_samples

  cell_references <- assigned_samples |>
    filter(.data$VALID_TMA, !is.na(.data$SECTOR)) |>
    summarise(
      N = dplyr::n(),
      GANP_P20 = as.numeric(stats::quantile(.data$TMA_TIME, probs = 0.20, names = FALSE)),
      TYPICAL_P50 = stats::median(.data$TMA_TIME),
      .by = c("RANGE_NM", "SECTOR", "RWY", "CLASS")
    )

  boundary_pairs <- proposal$cut_audit |>
    left_join(
      proposal$sector_definitions |>
        transmute(
          RANGE_NM = .data$RANGE_NM,
          CUT_ROUNDED = .data$BEARING_TO,
          LEFT_SECTOR = .data$SECTOR
        ),
      by = c("RANGE_NM", "CUT_ROUNDED")
    ) |>
    transmute(
      ICAO = airport,
      RANGE_NM = .data$RANGE_NM,
      CUT = .data$CUT_ROUNDED,
      PEAK_SEPARATION_DEG = .data$PEAK_SEPARATION_DEG,
      VALLEY_TO_LOWER_PEAK_RATIO = .data$VALLEY_TO_LOWER_PEAK_RATIO,
      PAIRWISE_SEPARATION = .data$PAIRWISE_SEPARATION,
      LEFT_SECTOR = .data$LEFT_SECTOR,
      RIGHT_SECTOR = .data$SECTOR
    )

  boundary_pairs |>
    left_join(
      cell_references |>
        rename(
          LEFT_SECTOR = .data$SECTOR,
          N_LEFT = .data$N,
          P20_LEFT = .data$GANP_P20,
          P50_LEFT = .data$TYPICAL_P50
        ),
      by = c("RANGE_NM", "LEFT_SECTOR")
    ) |>
    inner_join(
      cell_references |>
        rename(
          RIGHT_SECTOR = .data$SECTOR,
          N_RIGHT = .data$N,
          P20_RIGHT = .data$GANP_P20,
          P50_RIGHT = .data$TYPICAL_P50
        ),
      by = c("RANGE_NM", "RIGHT_SECTOR", "RWY", "CLASS")
    ) |>
    summarise(
      N_SHARED_CELLS = dplyr::n(),
      MIN_SHARED_CELL_N = min(pmin(.data$N_LEFT, .data$N_RIGHT)),
      WEIGHTED_ABS_P20_DIFF = stats::weighted.mean(
        abs(.data$P20_LEFT - .data$P20_RIGHT),
        w = pmin(.data$N_LEFT, .data$N_RIGHT)
      ),
      WEIGHTED_ABS_P50_DIFF = stats::weighted.mean(
        abs(.data$P50_LEFT - .data$P50_RIGHT),
        w = pmin(.data$N_LEFT, .data$N_RIGHT)
      ),
      .by = c(
        "ICAO", "RANGE_NM", "CUT", "PEAK_SEPARATION_DEG",
        "VALLEY_TO_LOWER_PEAK_RATIO", "PAIRWISE_SEPARATION"
      )
    )
}

study_arrival_results <- purrr::map(
  available_study_airports,
  function(airport) {
    airport_samples <- prepare_apdf_tma_reference_input_from_zip(
      zipped_archive_path = apdf_archive,
      files = paste0(airport, "_01-JAN-2024_31-DEC-2024.gz.parquet"),
      type = "parquet",
      ranges = c(40, 100)
    )
    density <- prepare_tma_bearing_density(
      tma_samples = airport_samples,
      airport = airport,
      phase = "ARR",
      ranges = c(40, 100),
      smoothing_bandwidth = smoothing_bandwidth
    )
    extrema <- identify_tma_bearing_extrema(
      tma_density = density,
      min_relative_prominence = minimum_relative_prominence
    )
    proposal <- propose_tma_sector_definitions(
      tma_density = density,
      extrema = extrema,
      rounding_increment = rounding_increment,
      valley_safety_fraction = valley_safety_fraction
    )
    multiscale_cuts <- purrr::map_dfr(
      analysis_bandwidths,
      function(bandwidth) {
        scale_density <- prepare_tma_bearing_density(
          tma_samples = airport_samples,
          airport = airport,
          phase = "ARR",
          ranges = c(40, 100),
          smoothing_bandwidth = bandwidth
        )
        scale_extrema <- identify_tma_bearing_extrema(
          tma_density = scale_density,
          min_relative_prominence = minimum_relative_prominence
        )
        mutate(
          propose_tma_sector_definitions(
            tma_density = scale_density,
            extrema = scale_extrema,
            # Retain raw-scale detail here; final proposals remain rounded to 5 degrees.
            rounding_increment = 1,
            valley_safety_fraction = valley_safety_fraction
          )$cut_audit,
          SMOOTHING_BANDWIDTH = bandwidth
        )
      }
    )

    list(
      density = density,
      proposal = proposal,
      support = summarise_tma_sector_support(airport_samples, proposal$sector_definitions),
      multiscale_cuts = multiscale_cuts,
      boundary_utility = build_boundary_utility(airport_samples, proposal, airport)
    )
  }
) |>
  stats::setNames(available_study_airports)

study_density <- purrr::imap_dfr(
  study_arrival_results,
  function(result, airport) mutate(result$density, ICAO = airport)
)
study_cuts <- purrr::imap_dfr(
  study_arrival_results,
  function(result, airport) mutate(result$proposal$cut_audit, ICAO = airport)
)
study_definitions <- purrr::imap_dfr(
  study_arrival_results,
  function(result, airport) mutate(result$proposal$sector_definitions, ICAO = airport)
)
study_multiscale_cuts <- purrr::imap_dfr(
  study_arrival_results,
  function(result, airport) mutate(result$multiscale_cuts, ICAO = airport)
)
study_boundary_utility <- purrr::map_dfr(study_arrival_results, "boundary_utility")

study_cut_stability <- study_cuts |>
  transmute(ICAO, RANGE_NM, CUT = .data$CUT_ROUNDED) |>
  purrr::pmap_dfr(function(ICAO, RANGE_NM, CUT) {
    airport_code <- ICAO
    range_nm <- RANGE_NM
    candidate_cut <- CUT
    tibble::tibble(
      ICAO = airport_code,
      RANGE_NM = range_nm,
      CUT = candidate_cut,
      N_STABLE_BANDWIDTHS = sum(purrr::map_lgl(
        analysis_bandwidths,
        function(bandwidth) {
          candidates <- study_multiscale_cuts |>
            filter(
              .data$ICAO == airport_code,
              .data$RANGE_NM == range_nm,
              .data$SMOOTHING_BANDWIDTH == bandwidth
            ) |>
            pull(.data$CUT_ROUNDED)
          any(circular_distance(candidates, candidate_cut) <= rounding_increment)
        }
      ))
    )
  })

study_boundary_register <- study_boundary_utility |>
  left_join(study_cut_stability, by = c("ICAO", "RANGE_NM", "CUT")) |>
  mutate(
    REVIEW_REASON = case_when(
      .data$PAIRWISE_SEPARATION < 0.75 ~ "Shallow valley",
      .data$PEAK_SEPARATION_DEG < 3 * smoothing_bandwidth ~ "Close peaks",
      .data$N_STABLE_BANDWIDTHS < length(analysis_bandwidths) ~ "Smoothing-sensitive cut",
      .default = NA_character_
    )
  ) |>
  arrange(.data$ICAO, .data$RANGE_NM, .data$CUT)

study_proposal_register <- study_definitions |>
  summarise(
    N_SECTORS = dplyr::n(),
    PROPOSED_SECTORS = paste(.data$SECTOR, collapse = "; "),
    NORTH_OVERRUN = first(.data$NORTH_OVERRUN),
    .by = c("ICAO", "RANGE_NM")
  ) |>
  left_join(
    study_boundary_register |>
      filter(!is.na(.data$REVIEW_REASON)) |>
      summarise(
        REVIEW_CUTS = paste0(.data$CUT, " (", .data$REVIEW_REASON, ")", collapse = "; "),
        .by = c("ICAO", "RANGE_NM")
      ),
    by = c("ICAO", "RANGE_NM")
  ) |>
  mutate(
    REVIEW_CUTS = coalesce(.data$REVIEW_CUTS, "None"),
    PROPOSAL_STATUS = case_when(
      .data$REVIEW_CUTS != "None" ~ "Analyst boundary decision required",
      .data$N_SECTORS > 6 ~ "Analyst complexity review",
      .default = "Initial proposal ready for approval"
    )
  ) |>
  arrange(.data$ICAO, .data$RANGE_NM)

study_data_availability <- tibble(
  ICAO = study_airports,
  DATA_STATUS = if_else(
    study_airports %in% available_study_airports,
    "2024 APDF analysed",
    "2024 APDF extract unavailable"
  )
)
```

```{r}
#| label: full-study-arrival-density-proposals
#| fig-width: 12
#| fig-height: 24
#| fig-cap: "2024 ARR density and proposed candidate sector boundaries for the 11 study airports available in the APDF archive. Each row is an airport; the columns are the 40 NM and 100 NM crossing horizons. Dashed lines require analyst approval before reference-time generation."

ggplot(
  study_density,
  aes(x = .data$BEARING_MID, y = .data$SMOOTHED_N)
) +
  geom_line(colour = "#2b6a8a", linewidth = 0.55) +
  geom_vline(
    data = study_cuts,
    aes(xintercept = .data$CUT_ROUNDED),
    colour = "#c8553d",
    linewidth = 0.45,
    linetype = "dashed",
    inherit.aes = FALSE
  ) +
  facet_grid(ICAO ~ RANGE_NM, scales = "free_y", labeller = label_both) +
  scale_x_continuous(breaks = seq(0, 360, by = 60)) +
  coord_cartesian(xlim = c(0, 360), expand = FALSE) +
  labs(
    title = "Study-airport ARR candidate sector boundaries, 2024",
    subtitle = paste(
      "Six-degree smoother;", rounding_increment,
      "degree valley-safe rounding; candidate boundaries only"
    ),
    x = "Bearing from aerodrome reference point (degrees)",
    y = "Smoothed movements per degree"
  ) +
  theme_minimal(base_size = 9) +
  theme(
    panel.grid.minor = element_blank(),
    strip.text.y = element_text(angle = 0)
  )
```

```{r}
#| label: full-study-arrival-proposal-register

study_proposal_register |>
  transmute(
    Airport = .data$ICAO,
    `Range (NM)` = .data$RANGE_NM,
    `Proposed sectors` = .data$PROPOSED_SECTORS,
    `North overrun` = .data$NORTH_OVERRUN,
    `Boundaries for review` = .data$REVIEW_CUTS,
    Status = .data$PROPOSAL_STATUS
  ) |>
  knitr::kable(format = "html")
```

```{r}
#| label: full-study-arrival-boundary-decisions

study_boundary_register |>
  filter(!is.na(.data$REVIEW_REASON)) |>
  transmute(
    Airport = .data$ICAO,
    `Range (NM)` = .data$RANGE_NM,
    `Candidate cut` = .data$CUT,
    `Review reason` = .data$REVIEW_REASON,
    `Peak gap (deg)` = round(.data$PEAK_SEPARATION_DEG, 1),
    `Valley / lower peak` = scales::percent(.data$VALLEY_TO_LOWER_PEAK_RATIO, accuracy = 0.1),
    `Stable bandwidths` = .data$N_STABLE_BANDWIDTHS,
    `Shared cells` = .data$N_SHARED_CELLS,
    `Weighted |P20 diff| (min)` = round(.data$WEIGHTED_ABS_P20_DIFF, 2),
    `Weighted |P50 diff| (min)` = round(.data$WEIGHTED_ABS_P50_DIFF, 2)
  ) |>
  knitr::kable(format = "html")
```

```{r}
#| label: full-study-arrival-data-availability

study_data_availability |>
  knitr::kable(format = "html")
```
